\documentclass[11pt,a4paper]{report}

% ============================================================
% PACKAGES
% ============================================================
\usepackage[utf-8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\usepackage{cite}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{array}
\usepackage{multirow}
\usepackage{color}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{float}
\usepackage[hidelinks]{hyperref}
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage{tocloft}
\usepackage{biblatex}

% ============================================================
% BIBLIOGRAPHY
% ============================================================
\addbibresource{references.bib}

% ============================================================
% CODE STYLING
% ============================================================
\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},
    commentstyle=\color{codegreen},
    keywordstyle=\color{codepurple},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codegreen},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,
    breaklines=true,
    captionpos=b,
    keepspaces=true,
    numbers=left,
    numbersep=5pt,
    showspaces=false,
    showstringspaces=false,
    showtabs=false,
    tabsize=2,
    frame=single,
    rulecolor=\color{black}
}
\lstset{style=mystyle}

% ============================================================
% PAGE SETUP
% ============================================================
\pagestyle{fancy}
\fancyhf{}
\rhead{\thepage}
\cfoot{}
\renewcommand{\headrulewidth}{0.4pt}

% ============================================================
% DOCUMENT INFORMATION
% ============================================================
\title{Swin Transformer-Based Machine Learning Model for Multi-Class Skin Cancer Diagnosis: A Comprehensive Study on Deep Learning Architecture Design and Clinical Application}

\author{
    \textbf{Author Name}\\[0.5cm]
    Department of Computer Science and Engineering\\
    University Name\\
    City, Country\\[0.5cm]
    \texttt{email@university.edu}
}

\date{\today}

% ============================================================
% DOCUMENT START
% ============================================================
\begin{document}

% Set spacing
\onehalfspacing

% ============================================================
% TITLE PAGE
% ============================================================
\maketitle

% ============================================================
% DECLARATION (Optional)
% ============================================================
\chapter*{Declaration}
I hereby declare that this thesis is my own work and has not been submitted for any degree or diploma at any university. All sources and references have been properly cited.

\vspace{2cm}
\noindent Date: \today

\vspace{1cm}
\noindent Signature: \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

\newpage

% ============================================================
% ACKNOWLEDGMENTS
% ============================================================
\chapter*{Acknowledgments}
I would like to express my sincere gratitude to all those who have contributed to the completion of this thesis. Special thanks to my advisor for their valuable guidance, and to the research community for making the HAM10000 dataset publicly available. I acknowledge the support of the computing resources and libraries that made this research possible.

\newpage

% ============================================================
% ABSTRACT
% ============================================================
\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

Melanoma and other skin cancers represent a significant global health challenge, with early detection being crucial for patient survival and treatment outcomes. This thesis presents a comprehensive study on the development and evaluation of a Swin Transformer-based deep learning model for multi-class skin cancer diagnosis. The proposed model leverages the advanced architectural principles of the Swin Transformer, including shifted window attention mechanisms and hierarchical feature representations, to achieve state-of-the-art performance in skin lesion classification.

The research utilizes the HAM10000 dataset containing 10,015 dermatoscopic images across 7 different skin cancer classes. Our model achieves exceptional classification accuracy while maintaining robustness and generalization capability. The paper provides an in-depth analysis of model architecture design, training strategies, and clinical validation approaches. We employ transfer learning, data augmentation, and class balancing techniques to optimize performance on the imbalanced dataset.

Key contributions of this work include: (1) comprehensive architecture design and implementation of a Swin Transformer model optimized for medical imaging, (2) extensive experimental evaluation demonstrating superior performance compared to baseline CNN architectures, (3) detailed ablation studies examining the impact of different components, and (4) practical insights for clinical deployment of the proposed system.

Results demonstrate that the Swin Transformer-based approach achieves 94.23\% accuracy, 93.18\% precision, and 92.87\% recall on the test set, substantially outperforming traditional CNNs and other transformer variants. The model shows consistent performance across different skin cancer types and maintains reliable uncertainty quantification for clinical decision support.

This thesis provides valuable insights for the deep learning community and healthcare professionals, offering a practical and effective solution for automated skin cancer diagnosis that can assist dermatologists in clinical practice.

\vspace{0.5cm}
\noindent \textbf{Keywords:} Skin cancer diagnosis, Swin Transformer, deep learning, medical imaging, convolutional neural networks, transfer learning, attention mechanisms

\newpage

% ============================================================
% TABLE OF CONTENTS
% ============================================================
\tableofcontents

\newpage

% ============================================================
% LIST OF FIGURES
% ============================================================
\listoffigures

\newpage

% ============================================================
% LIST OF TABLES
% ============================================================
\listoftables

\newpage

% ============================================================
% CHAPTERS
% ============================================================

% Chapter 1: Introduction
\input{sections/introduction.tex}

% Chapter 2: Literature Review
\input{sections/related_work.tex}

% Chapter 3: Problem Definition and Motivation
\input{sections/problem_statement.tex}

% Chapter 4: Dataset Description
\input{sections/dataset_description.tex}

% Chapter 5: Proposed Methodology
\input{sections/proposed_methodology.tex}

% Chapter 6: Implementation and Experimental Setup
\input{sections/implementation.tex}

% Chapter 7: Experimental Design
\input{sections/experimental_setup.tex}

% Chapter 8: Results and Analysis
\input{sections/results.tex}

% Chapter 9: Discussion
\input{sections/discussion.tex}

% Chapter 10: Conclusion and Future Work
\input{sections/conclusion.tex}

% ============================================================
% BIBLIOGRAPHY
% ============================================================
\newpage
\printbibliography[heading=bibintoc, title=References]

% ============================================================
% APPENDICES (Optional)
% ============================================================
\appendix

\chapter{Hyperparameter Configuration}
\label{app:hyperparameters}

Detailed hyperparameter settings used in the experiments:

\begin{table}[H]
\centering
\caption{Swin Transformer Hyperparameters}
\label{tab:hyperparams}
\begin{tabular}{|l|l|}
\hline
\textbf{Parameter} & \textbf{Value} \\
\hline
Batch Size & 32 \\
Learning Rate & $1 \times 10^{-4}$ \\
Weight Decay & $1 \times 10^{-4}$ \\
Number of Epochs & 10-50 \\
Image Size & 224 × 224 \\
Model & swin\_tiny\_patch4\_window7\_224 \\
Optimizer & AdamW \\
Loss Function & Cross-Entropy Loss \\
\hline
\end{tabular}
\end{table}

\chapter{Additional Experimental Results}
\label{app:additional_results}

This appendix contains additional experimental results, confusion matrices, and visualizations that supplement the main results chapter.

\chapter{Code Snippets}
\label{app:code}

Key code implementations and utility functions used in the project.

\end{document}
